---
title: "Data Manipulation Worksheet"
output: html_notebook
---


# Get started

It is a good habit to first create a RStudio project. Go to the menu, choose `File`, then `New Project`, to initiate a new project. Once we have RStudio project set up, we need to copy the data files and place them in the project directory, preferably under the folder named `data` within the project folder.

The worksheet assumes that you have done all of the above steps!

# The first step

We will load all the necessary libraries into R session. Since all packages we will use in this worksheet are parts of the `tidyverse`, we will use it.

```{r}
library(tidyverse)
```

Next, we read all the data files.

```{r}
flights <- read_csv("data/flights.csv")
weather <- read_csv("data/weather.csv")
planes <- read_csv("data/planes.csv")
airports <- read_csv("data/airports.csv")
```

# Your turn questions

1. Find all flights:
  a) To SFO or OAK
  b) In January
  c) Delayed by more than an hour
  d) That departed between midnight and 5 a.m.
  e) Where the arrival delay was more than twice the departure delay
  
```{r}
# Your code goes here

```

2. Use `select()`
  a) Which flights were most delayed?
  b) Which flights caught up the most time during the flight?
  
```{r}
# Your code goes here

```

3. Use `mutate()`
  a). Add a new variable that shows how much time was made up or lost in flight 
  b) Compute speed in mph from time (in minutes) and distance (in miles). Which flight flew the fastest?

```{r}
# Your code goes here

```

4. How might you summarise `dep_delay` for each day?

```{r}
# Your code goes here

```

5. Crete data pipelines to answer the following questions:
  a) Which destinations have the highest average delays?
  b) Which flights (i.e., carrier + flight) happen every day? Where do they fly to?
  
```{r}
# Your code goes here

```

6. What weather conditions are associated with delays leaving in Houston? Use graphics to explore.

```{r}
hourly_delay <- flights %>% mutate(date = as_date(date)) %>%
  group_by(date, hour) %>%
  filter(!is.na(dep_delay)) %>%
  summarise(
    delay = mean(dep_delay),
    n = n()
  ) %>%
  filter(n > 10)

delay_weather <- hourly_delay %>% left_join(weather)

# You code goes below this comment

```
