---
title: "Data Visualisation with ggplot 2"
date: "August 24, 2018"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# Introduction

> "The simple graph has brought more information to the data analyst’s mind than any other device.” — John Tukey

- One of the biggest benefits of using R is its graphical capacity.
- The package `ggplot2` and its related packages create the most elegant and most versatile graphs.
- `ggplot2` implements the **grammar of graphics**, a coherent system for describing and building graphs
- To learn more, read "The Layered Grammar of Graphics", [http://vita.had.co.nz/papers/layered-grammar.pdf]( http://vita.had.co.nz/papers/layered-grammar.pdf)

---
# First example

**Question:** Do cars with big engines use more fuel than cars with small engines?

--

```{r, message=FALSE}
library(tidyverse)
data(mpg)
```

*Checkpoint:* How many rows are in `mpg`? How many columns?

--

Variables of interest: `displ` (engine size) and `hwy` (fuel efficientcy)

---
# First `ggplot2` graph

```{r, fig.height=5}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

---
# A graphing template

```
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

To make a graph, replace the bracketed sections  with a data set, a *geom* function, or a collection of *mappings*.

--

*Check points:*
- Make a scatterplot of `hwy` versus `cyl`.
- What happens if you make a scatterplot if `class` versus `drv`? (You can look up the help documnet of the data set `mpg` for definition of `class` and `drv`)

---
# Aesthetic mappings

```{r, echo=FALSE, fig.height=4}
outliers <- mpg %>% filter(hwy > 20, displ > 5)
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy)) +
  geom_point(aes(x = displ, y = hwy), data = outliers, color = "red")
```

One group of points (highlighted in red) seems to fall outside of the linear trend. These cars have a higher mileage than you might expect. How can you explain these cars?

--

*The type of the cars might provide more insight!!*

---
# Aesthetic mappings

- An aesthetic is a visual property of the objects in a graph
- Aesthetics include the size, the shape, the color of points, etc.

--
```{r, fig.height=4}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

---
# In-class activities

1. Use different aesthetics for `class`. Your options are: color, size, shape, alpha. Can you say something about those outliers?

--

2. Is this possible to use multiple aesthetics on the same graph? Give an example.

---
# Aesthetics

You can also set the aesthetic properties of your geom manually. 
```{r, fig.height=3}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

--

**Quick question:** What’s gone wrong with this code? Why are the points not blue?

```{r, eval=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

---
# Facets

- Another way to add variables (esp., categorical) is to split the plot into **facets**, subplots that each display one subset of the data.

```{r, fig.height=4}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

---
# Facets

```{r, fig.height=4}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

--

*Quick question:* Why do the empty cells in plot mean?

---
# Geometric objects

- A **geom** is the geometrical object that a plot uses to represent data
- Bar charts use *bar geoms*, line charts use *line geoms*, boxplots use *boxplot geoms*, and so on.

--

.pull-left[
```{r, fig.height=4}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```
]

.pull-right[
```{r, , fig.height=4, message=FALSE}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```
]

---
# Geometric objects

- Every geom functions takes a `mapping` argument.

- Not every aesthetic works with every geom

--

- `geom_smooth()` does not have "shape" aesthetic (it does have line type).

--

```{r, fig.height=4, message=FALSE}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

---
# Geometric objects

- We can have multiple geoms in the same plot:

```{r, fig.height=4, message=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

---

# Geometric objects

```{r, fig.height=4, eval=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + #<<
  geom_smooth(mapping = aes(x = displ, y = hwy)) #<<
```

--

We can reduce the duplication as:

```{r, eval=FALSE, message=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() +
  geom_smooth()
```

`ggplot2` treats these mappings as global mappings that share among geom objects.
---
# Geometric objects

- Local mappings extends global mappings

```{r, fig.height=4, message=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) +
  geom_smooth() 
```

---
# Geometric objects

- Local mappings overrides global mappings

```{r, fig.height=4, message=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) +
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE) 
```

---
# Labels in `ggplot2`

- The labels can be done via `labs()`

```{r, fig.height=4, message=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) +
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE) +
  labs(title = "Engine size vs fuel efficiency",
       x = "Engine size (liters)",
       y = "Highway fuel efficiency (mpg)")
```

---
# Final words

- So far, we just visited a tip of an iceberg.
- Many amazing graphs can be created with a few lines of code.
- Check out the *R graphics gallery* at [https://www.r-graph-gallery.com/](https://www.r-graph-gallery.com/)

```{r, eval=FALSE, message=FALSE, warning=FALSE}
library(maps)
states <- map_data("state")
top100 <- world.cities %>% filter(country.etc=="USA") %>%
  mutate(pop_in_mil = pop/1000000) %>%
  arrange(desc(pop_in_mil)) %>%
  slice(1:100)

ggplot() + 
  geom_polygon(data=states, aes(x=long, y=lat, group = group), 
               color="black", alpha=0.3) +
  geom_point(data=top100,mapping=aes(x=long,y=lat,size=pop_in_mil))+
  coord_map() + xlim(-130, -70) + ylim(25, 50) +
  labs(title = "Top 100 biggest cities in US ")
```
---
class: center, middle

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=10}
library(maps)
states <- map_data("state")
top100 <- world.cities %>% filter(country.etc=="USA") %>%
  mutate(pop_in_mil = pop/1000000) %>%
  arrange(desc(pop_in_mil)) %>%
  slice(1:100)
ggplot() + 
  geom_polygon(data=states, aes(x=long, y=lat, group = group), color="black", alpha=0.3) +      geom_point(data=top100, mapping=aes(x=long, y=lat, size = pop_in_mil)) +
  coord_map() + xlim(-130, -70) + ylim(25, 50) +
  labs(title = "Top 100 biggest cities in US ")
```



